---
title: "MovieLens Dataset"
author: Nicola Gennaro
date: September 2018
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache=TRUE)
```



## MovieLens Dataset

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

library(tidyverse)
library(dplyr)
library(ggplot2)
library(reshape2)
library(knitr)
library(kableExtra)
library(cowplot)

options(knitr.table.format = "html")
```

<b>MovieLens</b> is a web-based recommender system and virtual community that recommends movies for its users to watch, based on their film preferences using collaborative filtering of members' movie ratings and movie reviews.

The dataset contains:

- over 1.000.000 ratings 
- from more than 6.000 users
- about 4.000 movies

It's divided in three datasets: Users, Movies, Ratings





## Users Dataset

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
users <- read.csv("/home/nicola/DSSC/Database/project/MovieLens/dataset/users.csv")
users$age <- factor(users$age, levels = c("Under 18", "18-24", "25-34", "35-44", "45-49", "50-55", "56+"))
levels(users$age) <- c("<18", "18-24", "25-34", "35-44", "45-49", "50-55", "56+")
```


```{r}
users[1:10,] %>% kable(align = "c") %>%
  kable_styling(bootstrap_options = "striped", font_size = 15, full_width = F)
```



## Male-Female Users

Male Proportion: 0.717

```{r}
ggplot(data=users) + 
  geom_bar(mapping = aes(x=gender, fill=gender)) +
  xlab("Gender") +
  ylab("Count") +
  labs(title="Gender Counts")
```



## Users' Ages

The age is divided in 7 categories:

```{r}
ggplot(data=users) + 
  geom_bar(mapping = aes(x=age, fill=age)) +
  xlab("Age") +
  ylab("Count") +
  labs(title="Age Counts")
```



## Occupations

There are 20 diferent occupations

```{r}
ggplot(data=users) + 
  geom_bar(mapping = aes(x=occupation, fill=gender)) +
  xlab("Occupation") +
  ylab("Count") +
  labs(title="Occupation Counts MF") +
  coord_flip() 
```


## Occupation-Gender Proportions

```{r, echo=FALSE}
fm <- sum(users$gender == "M")/length(users$gender)
```

```{r}
ggplot(data=users) + 
  geom_bar(mapping = aes(x=occupation, fill=gender), position="fill") +
  geom_hline(aes(yintercept=fm)) + 
  xlab("Occupation") +
  ylab("Count") +
  labs(title="Occupation Proportion - Gender") +
  coord_flip()
```


## Occupation-Age Proportions

```{r}
ggplot(data=users) + 
  geom_bar(mapping = aes(x=occupation, fill=age), position="fill") +
  xlab("Occupation") +
  ylab("Proportion") +
  labs(title="Occupation Proportion - Age") +
  coord_flip()
```




<!-- ZIP -->

## ZipCode Dataset

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
zip <- read.csv("/home/nicola/DSSC/Database/project/MovieLens/dataset/zip.csv")

zip1 <- zip[ c("Zipcode", "City", "State", "Lat", "Long", "Country") ]
colnames(zip1) <- c("zipCode", "city", "state", "lat", "long", "country")
zip1 <- zip1[ !duplicated(zip1$zipCode), ]

use.zip <- inner_join(users, zip1, by="zipCode" )

us.user <- use.zip[ use.zip$lat > 24 &  use.zip$lat < 50, ]
us.user <- us.user[ complete.cases(us.user), ]

states <- c("CA", "MN", "TX", "NY")

new_states <- rep("other", dim(us.user)[1])

for( s in states){
  new_states[ us.user$state == s ] <- s
}

new_states <- factor(new_states, levels = c(states, "other"))

states_colors <- c("CA" = "yellow", "MN" = "red", "TX" = "blue", "NY" = "green",   "other" = "black")

states_df <- us.user[ us.user$state %in% states, ]
```

```{r}
ggplot(data=us.user) +
  geom_point(mapping = aes(x = long, y = lat)) + coord_fixed(ratio=1.5) +
  xlab("") +
  ylab("") +
  labs(title="USA users") +
  theme_bw()
```



<!-- ## States Analysis -->


<!-- ```{r} -->
<!-- p1 <- ggplot(data=states_df) +  -->
<!--   geom_bar(mapping = aes(x=state, fill=state)) + -->
<!--   xlab("State") + -->
<!--   ylab("") + -->
<!--   theme_bw() + -->
<!--   labs(title="State Users") + -->
<!--   scale_fill_manual(values = states_colors) -->

<!-- p2 <- ggplot(data=states_df) +  -->
<!--   geom_bar(mapping = aes(x=occupation, fill=state), position="fill") + -->
<!--   xlab("Occupation") + -->
<!--   ylab("") + -->
<!--   theme_bw() + -->
<!--   labs(title="State Occupations Proportions") + -->
<!--   scale_fill_manual(values = states_colors) + -->
<!--   coord_flip() -->


<!-- plot_grid(p1, p2, align="o", ncol=2, rel_widths = c(1/3, 2/3)) -->
<!-- ``` -->




## Movies Dataset

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
movies <- read.csv("/home/nicola/DSSC/Database/project/MovieLens/dataset/movies.csv")
movies <- movies[ complete.cases(movies), ]
```

```{r}
movies[1:10, c(1,2,3,5,6,7)] %>% kable(align = "c") %>%
  kable_styling(bootstrap_options = "striped", font_size = 15, full_width = F)
```


## Movies Years

```{r}
ggplot(data=movies) + 
  geom_bar(mapping=aes(x=year)) +
  xlab("Year") +
  ylab("Count") +
  labs(title="Movies Year Counts")
```

## Movies Genres

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
summ <- colSums(movies[, 5:22])
summdf <- data.frame(summ)
summdf$genre <- names(summ)
```

```{r}
ggplot(data=summdf) + 
  geom_bar(mapping = aes(x=genre, y=summ), stat="identity") +
  xlab("Genre") +
  ylab("Count") +
  labs(title="Movies Genres Counts") +
  coord_flip()
```



## Ratings Dataset


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
ratings <- read.csv("/home/nicola/DSSC/Database/project/MovieLens/dataset/ratings.csv")
```


```{r}
ratings[1:10, ] %>% kable(align = "c") %>%
  kable_styling(bootstrap_options = "striped", font_size = 15, full_width = F)
```


## Ratings

People tend to see and rate movies they like

```{r}
ratings$Rating <- factor(ratings$rating, levels = c(1,2,3,4,5))

p1 <- ggplot(data=ratings) + 
  geom_bar(mapping = aes(x=Rating, fill=Rating)) +
  xlab("Rating") +
  ylab("Count") +
  labs(title="Ratings Counts")

ratings$Rating <- NULL

p2 <- ggplot(data=ratings, aes(x="", y=rating)) +
  stat_summary(fun.data="mean_sdl", fun.args = list(mult=1), geom="crossbar", width=0.5)+ coord_fixed(ratio=2.5) + xlab("Rating") + ylab("Mean") +
  labs(title="Mean Rating")

?plot_grid
plot_grid(p1, p2, align="o", ncol=2, rel_widths = c(2/3, 1/3))
#grid.arrange(p1, p2, ncol=2)
```


## Male-Female Ratings

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rat.use <- inner_join(ratings, users, by="userID")
fm.rat <- sum(rat.use$gender == "M")/length(rat.use$gender)
```

```{r}
ggplot(data=rat.use) + 
  geom_bar(mapping = aes(x=rating, fill=gender), position="fill") +
  geom_hline(aes(yintercept=fm.rat), linetype=3) +
  geom_hline(aes(yintercept=fm)) +
  xlab("Rating") +
  ylab("MF Proportion") +
  labs(title="Ratings Counts M-F")
```



## Age Ratings

Older people's ratings are higher then younger ones...

```{r}
ggplot(data=rat.use) + 
  theme_bw() + 
  geom_bar(mapping = aes(x=rating, fill=age), position="fill") +
  xlab("Rating") +
  ylab("Age Proportion") +
  labs(title="Ratings Counts Age")
```


## Age - Year

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
rat.use.year <- inner_join(rat.use, movies, by="movieID")
```


```{r}
ggplot(data=rat.use.year, mapping = aes(x = age, y = year, color=age)) +
  geom_boxplot() +
  xlab("Age") +
  ylab("Year") +
  labs(title="Mean year vs Age")
```




## Decade - Age

The tendency of older people to give higher votes seems to extend to all decades.
```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
decade <- function(x){
  return(x %/% 10 * 10 )
}

rat.use.year <- rat.use.year %>% mutate(decade = decade(year))
rat.use.year$decade <- factor(rat.use.year$decade, levels = c("1910", "1920", "1930", "1940", "1950", "1960", "1970", "1980", "1990", "2000"))
```


```{r}
ggplot(data=rat.use.year, mapping = aes(x = age, y = rating, col=age)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  stat_summary(fun.data="mean_sdl", fun.args = list(mult=1), geom="crossbar", width=0.5) + 
  facet_grid(~decade) +
  xlab("Age") +
  ylab("Mean Rating") +
  labs(title="Mean rating for each Age and Decade")
```



## Age - Decade

But from this plot is clear that for all ages, older movies have a higher vote

```{r}
ggplot(data=rat.use.year, mapping = aes(x = decade, y = rating, col=decade)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  stat_summary(fun.data="mean_sdl", fun.args = list(mult=1), geom="crossbar", width=0.5) + 
  facet_grid(~age) +
  xlab("Decade") +
  ylab("Mean Rating") +
  labs(title="Mean rating for each Age and Decade")
```




## Genre - Gender

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
create_df <- function(dataframe, extract, vec, genres){
  mf <- data.frame()
  for( v in vec ){
    for( genre in genres ){
      x <- extract(dataframe, v, genre)
      y <- data.frame(v, genre, x[1], x[2])
      mf <- rbind(mf, y)
    }
  }
  return(mf)
}

genres <- c("Drama", "Comedy", "Action", "Horror", "Western", "Romance", "Children.s", "Musical", "War")
```


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
genders <- c("F", "M")

extract_sex <- function(dataframe, sex, genre){
  df1 <- dataframe[ dataframe$gender == sex & dataframe[genre] == 1, "rating"]
  return(c(mean(df1), sd(df1)))
}

m <- create_df(rat.use.year, extract_sex, genders, genres)
colnames(m) <- c("gender", "genre", "mean", "sd")
```


```{r, message=FALSE, warning=FALSE}
ggplot(m, aes(x=gender, y=mean, col=gender)) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2) +
  geom_line() +
  geom_point() +
  theme_bw() +
  facet_grid(~genre) +
  xlab("Gender") +
  ylab("Mean Rating") +
  labs(title="Mean rating vs Gender")
```

## Genre - Gender

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
extract_count <- function(gender, genre){
  x <- count(rat.use.year[ rat.use.year$gender == gender & rat.use.year[genre] == 1, ])
  return(x)
}

d <- data.frame()
for( gender in genders){
  for( genre in genres){
    n <- extract_count(gender, genre)
    y <- data.frame(gender, genre, n)
    d <- rbind(d, y)
  }
}

colnames(d) <- c("gender", "genre", "nn")


fm.rat <- sum(rat.use$gender == "M")/length(rat.use$gender)
```

```{r}
ggplot(data=d, aes(x=genre)) + 
  geom_col(aes( y=nn, fill=gender ), position="fill") + 
  geom_hline(aes(yintercept=fm.rat)) + 
  xlab("Movie Genre") +
  ylab("Proportion") +
  labs(title="Ratings Counts - Gender")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 9))
```




## Genre - Age

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
extract_age <- function(dataframe, age, genre){
  df1 <- dataframe[ dataframe$age == age & dataframe[genre] == 1, "rating"]
  return(c(mean(df1), sd(df1)))
}

ages <- c("<18", "18-24", "25-34", "35-44", "45-49", "50-55", "56+")

m <- create_df(rat.use.year, extract_age, ages, genres)
colnames(m) <- c("age", "genre", "mean", "sd")
```


```{r, message=FALSE, warning=FALSE}
ggplot(m, aes(x=age, y=mean)) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2) +
  geom_line() +
  geom_point() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) +
  facet_grid(~genre) +
  xlab("Age") +
  ylab("Mean Rating") +
  labs(title="Mean rating for each Age and Genre")
```


## Genre - Age

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
extract_count <- function(age, genre){
  x <- count(rat.use.year[ rat.use.year$age == age & rat.use.year[genre] == 1, ])
  return(x)
}

d <- data.frame()
for( age in ages){
  for( genre in genres){
    n <- extract_count(age, genre)
    y <- data.frame(age, genre, n)
    d <- rbind(d, y)
  }
}

colnames(d) <- c("age", "genre", "nn")
```

```{r}
ggplot(data=d, aes(x=genre)) + 
  geom_col(aes( y=nn, fill=age ), position="fill") + 
  xlab("Movie Genre") +
  ylab("Proportion") +
  labs(title="Ratings Counts - Age")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 9))
```



<!-- ############################## -->

## Genre - Occupation

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
jobs <- c("K-12 student", "scientist", "programmer", "lawyer", "artist", "retired")

extract_job <- function(dataframe, job, genre){
  df1 <- dataframe[ dataframe$occupation == job & dataframe[genre] == 1, "rating"]
  return(c(mean(df1), sd(df1)))
}

m <- create_df(rat.use.year, extract_job, jobs, genres)
colnames(m) <- c("occupation", "genre", "mean", "sd")
```



```{r, message=FALSE, warning=FALSE}
ggplot(m, aes(x=occupation, y=mean)) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2) +
  geom_line() +
  geom_point() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7)) +
  facet_grid(~genre) +
  xlab("Occupation") +
  ylab("Mean Rating") +
  labs(title="Mean rating for each Occupation and Genre")
```


<!-- ## Genre - State -->

<!-- ```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE} -->
<!-- df <- inner_join(ratings, use.zip, by="userID") -->
<!-- df <- inner_join(movies, df, by="movieID") -->


<!-- extract_state <- function(dataframe, state, genre){ -->
<!--   df1 <- dataframe[ dataframe$state == state & dataframe[genre] == 1, "rating"] -->
<!--   return(c(mean(df1), sd(df1))) -->
<!-- } -->

<!-- m <- create_df(df, extract_state, states, genres) -->
<!-- colnames(m) <- c("state", "genre", "mean", "sd") -->
<!-- ``` -->

<!-- ```{r, message=FALSE, warning=FALSE} -->
<!-- ggplot(m, aes(x=state, y=mean)) + -->
<!--   geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.2) + -->
<!--   geom_line() + -->
<!--   geom_point() + -->
<!--   theme_bw() + -->
<!--   theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) + -->
<!--   facet_grid(~genre) + -->
<!--   xlab("State") + -->
<!--   ylab("Mean Rating") -->
<!-- ``` -->




## Conclusions

It's hard to predict the taste of a user using only his/her generic data:

- Age is slightly related to the year of rated movies and older users tend to rate older movies.
- In general older movies received higher votes, while older people give higher votes.
- Gender is slightly related to the movies' genre but only for some of them (especially Musicals and Childrens' movies)
- The occupation of a user contain no information about his/her tastes.